Performing graph-based clustering with rOpenScPCA

### 2024-12-13 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
rearrange calculating qc metrics on existing clusters

```{r cluster sce nondefault}
# Calculate clusters with non-default parameters
cluster_results_df <- rOpenScPCA::calculate_clusters(
  pca_matrix,
  algorithm = "leiden",
  nn = 15,
  objective_function = "modularity"
)

@@

# Now, pass in the same arguments customizing parameters here
stability_results_leiden <- rOpenScPCA::calculate_stability(
  pca_matrix,
  cluster_df_leiden,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

stability_results_leiden
```

## Calculate QC metrics on existing clusters

@@

In this section, we'll present examples of how you can calculate the silhouette width, neighborhood purity, and cluster stability from existing cluster assignments.

@@

To demonstrate this, we'll first create a Seurat object from our SCE object. 
We'll use the function `rOpenScPCA::sce_to_seurat()` for this, setting `use_symbols = FALSE` to preserve Ensembl ids as the feature names rather than converting to gene symbols.

```{r sce to seurat}
# Convert the SCE to a Seurat object using rOpenScPCA
seurat_obj <- rOpenScPCA::sce_to_seurat(sce, use_symbols = FALSE)

@@

The `rOpenScPCA::sce_to_seurat()` function will carry over the existing SCE column `cluster` into the Seurat object, which we'll use here for demonstration. 
If you calculate clusters within the Seurat ecosystem, you may need to use a different column name here, e.g. `seurat_clusters`.

```{r seurat cluster}
head(seurat_obj$cluster)
```

To calculate QC metrics on these clusters, we'll need two pieces of information:

- The PCA matrix
- A data frame with columns `cell_id` and `cluster`

We'll create those here:

```{r prepare seurat input}
# Extract the PCA matrix from the Seurat object
seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")

# Create a data frame for input
seurat_cluster_df <- data.frame(
  cell_id = colnames(seurat_obj),
  cluster = seurat_obj$cluster
)
```

Now, we can run `calculate_silhouette()` and `calculate_purity()`:

```{r seurat silhouette}
seurat_silhouette_df <- rOpenScPCA::calculate_silhouette(
  seurat_pca_matrix,
  seurat_cluster_df
)
head(seurat_silhouette_df)
```

```{r seurat purity}
seurat_purity_df <- rOpenScPCA::calculate_purity(
  seurat_pca_matrix,
  seurat_cluster_df
)
head(seurat_purity_df)
```

@@

scpca_cluster_df
```

We can run evaluation functions using this data frame and the PCA matrix from the object.

```{r scpca silhouette}
# Calculate silhouette width
scpca_silhouette_df <- rOpenScPCA::calculate_silhouette(
  pca_matrix,
  scpca_cluster_df
)
head(scpca_silhouette_df)
```

```{r scpca purity}
# Calculate neighborhood purity
scpca_purity_df <- rOpenScPCA::calculate_purity(
  pca_matrix,
  scpca_cluster_df
)
head(scpca_purity_df)
```

When running `rOpenScPCA::calculate_stability()`, we'll specify the same parameters originally used to build the clusters.
We only recommend evaluating clusters with `rOpenScPCA::calculate_stability()` if you know the original parameters used.

```{r scpca stability}
scpca_stability_df <- rOpenScPCA::calculate_stability(
  pca_matrix,
  scpca_cluster_df,
  # provide ScPCA clustering parameters
  algorithm = "louvain",
  weighting = "jaccard",
  nn = 20
)
scpca_stability_df
```

Instead of a matrix, you can alternatively pass in an SCE or Seurat object that contains a matrix.
We show examples of this below with `rOpenScPCA::calculate_clusters()`, but this will also work for any of the evaluation functions as well:

```{r run on seurat}
# Calculate clusters from a Seurat object using default parameters
cluster_results_df <- rOpenScPCA::calculate_clusters(seurat_obj)
cluster_results_df
```

`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects (as they were in the above examples).
If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.

>>>>

```{r cluster sce nondefault}
# Calculate clusters with non-default parameters
cluster_results_df <- rOpenScPCA::calculate_clusters(
  pca_matrix,
  algorithm = "leiden",
  nn = 15,
  objective_function = "modularity"
)
```

@@

# Now, pass in the same arguments customizing parameters here
stability_results_leiden <- rOpenScPCA::calculate_stability(
  pca_matrix,
  cluster_df_leiden,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)
```

Instead of a matrix, you can alternatively pass in an SCE or Seurat object that contains a matrix.

We show an example of this below with and SCE object and `rOpenScPCA::calculate_clusters()`, but this will also work for any of the evaluation functions as well and has the same syntax for Seurat objects.

`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects.
If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.

## Calculating QC metrics on existing clusters

@@

In this section, we'll present examples of how you can calculate the silhouette width, neighborhood purity, and cluster stability from existing cluster assignments within objects.

@@

To demonstrate this, we'll convert our SCE object to a Seurat using the function `rOpenScPCA::sce_to_seurat()`.
We'll set `use_symbols = FALSE` to preserve Ensembl ids as the feature names rather than converting to gene symbols.
Then, we'll use a simple Seurat pipeline to obtain clusters.

```{r sce to seurat, message = FALSE}
# Convert the SCE to a Seurat object using rOpenScPCA, preserving Ensembl ids as feature names
seurat_obj <- rOpenScPCA::sce_to_seurat(sce, use_symbols = FALSE)

# Calculate clusters with Seurat
seurat_obj <- seurat_obj |>
  SCTransform() |>
  RunPCA() |>
  FindNeighbors() |>
  FindClusters()

@@

To calculate QC metrics on these clusters, we'll need to create a data frame with columns `cell_id` and `cluster`:

```{r prepare seurat input}
# Create a data frame for input
seurat_cluster_df <- data.frame(
  cell_id = colnames(seurat_obj),
  cluster = seurat_obj$seurat_clusters
)

head(seurat_cluster_df)
```

Now, we can run `rOpenScPCA::calculate_silhouette()` and `rOpenScPCA::calculate_purity()` using this data frame and the Seurat object:

```{r seurat silhouette}
seurat_silhouette_df <- rOpenScPCA::calculate_silhouette(
  seurat_obj,
  seurat_cluster_df
)
```

```{r seurat purity}
seurat_purity_df <- rOpenScPCA::calculate_purity(
  seurat_obj,
  seurat_cluster_df
)
```

@@

head(scpca_cluster_df)
```

We can run evaluation functions using this data frame and the SCE object.

```{r scpca silhouette}
# Calculate silhouette width
scpca_silhouette_df <- rOpenScPCA::calculate_silhouette(
  sce,
  scpca_cluster_df
)
```

```{r scpca purity}
# Calculate neighborhood purity
scpca_purity_df <- rOpenScPCA::calculate_purity(
  sce,
  scpca_cluster_df
)
```

When running `rOpenScPCA::calculate_stability()`, we'll specify the same parameters originally used to build the clusters.
Generally speaking, we only recommend evaluating clusters with `rOpenScPCA::calculate_stability()` if you know the original parameters used.

```{r scpca stability}
scpca_stability_df <- rOpenScPCA::calculate_stability(
  sce,
  scpca_cluster_df,
  # provide ScPCA clustering parameters
  algorithm = "louvain",
  weighting = "jaccard",
  nn = 20
)
```
